---
title: "Connecting points in Normal Probability Plot"
output: 
html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

<div id="Overview">
We're going to connect all the points in a normal probability plot with lines. There will actually be one normal probability plot for each level of a categorical variable. **The dataset used is the <em>nc</em> dataset from Duke's Coursera's statsr package**
</div>

## Load packages
```{r}
library(dplyr)
library(magrittr)
library(statsr)
```

## Load data
```{r}
data(nc)
```

Here's a regular normal probability plot of the ***weight*** variable on nonmissing levels of the ***habit*** variable. By "regular" I mean that we're not connecting the points with lines. 

```{r using-points}
ncSmokers<-nc%>%filter(habit=="smoker")
ncNonsmokers<-nc%>%filter(habit=="nonsmoker")

par(mfrow=c(1,2))

qqnorm(ncSmokers$weight,main="NPP plot of weights of smokers\n in nc dataset")
qqline(ncSmokers$weight)

qqnorm(ncNonsmokers$weight,main="NPP plot of weights of\n nonsmokers in nc dataset")
qqline(ncNonsmokers$weight)
```

Now here's a normal probability plot of the ***weight*** variable on nonmissing levels of the habit variable wherein datapoints are connected by lines. What seems wrong with it?

```{r using-lines-wrong}
par(mfrow=c(1,2))

qqnorm(ncSmokers$weight,main="NPP plot of weights of smokers\n in nc dataset",type="l")
qqline(ncSmokers$weight)

qqnorm(ncNonsmokers$weight,main="NPP plot of weights of\n nonsmokers in nc dataset",type="l")
qqline(ncNonsmokers$weight)
```

**The plot doesn't <em>seem</em> to reflect the data. Here's what's going on. The lines connect points in the order that they appear in the dataset- i.e. the point is not necessarily going to be connected to its lexicographic successor- this is necessary for the plot to look as you want it. To make this so, we'll sort each subsample by weight before plotting it. **

```{r using-line-correct}
par(mfrow=c(1,2))

ncSmokers<-arrange(ncSmokers,weight)
qqnorm(ncSmokers$weight,main="NPP plot of weights of smokers\n in nc dataset",type="l")
qqline(ncSmokers$weight)

ncNonsmokers<-arrange(ncNonsmokers,weight)
qqnorm(ncNonsmokers$weight,main="NPP plot of weights of\n nonsmokers in nc dataset",type="l")
qqline(ncNonsmokers$weight)
```

**Ah- much better!**