---
title: "Some one-way ANOVA examples"
output:
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Load packages
```{r}
library(magrittr)
```
## Exercise 19.1 (source: "The Book of R"  by Tillman Davies)

***a./b. checks for normality of each group and constant variance across the groups***

Below we'll make a data frame that holds data for the depths of archaeological findings at 5 different sites. The depths are reported in centimeters.

```{r make-data-frame}
depths<-c(93,120,65,105,115,82,99,87,100,90,78,95,93,88,110,85,45,80,28,75,70,65,55,50,40,100,75,65,40,73,65,50,30,45,50,45,55,96,58,95,90,65,80,85,95,82)

siteNums<-rep(strsplit("Site1 Site2 Site3 Site4"," ")%>%unlist,times=c(15,10,12,9))

df<-data.frame(depth=depths,siteNum=siteNums)
```

Now let's check for normality of depths at each site and constant variance of depths across sites. We'll assume that depths at a particular site are independent of one another and that the depths at one of the four sites are independent of those at any other of the four sites.

```{r Check normality and constant variance using side by side boxplot }
boxplot(df$depth~df$siteNum,main="Side by side boxplot of\ndepths by site",ylab="depth (cm)")
siteMeans<-with(df,tapply(depth,siteNum,FUN=mean))
points(1:4,siteMeans,pch=4,cex=1.5)
```
The side-by-side boxplots indicate that each distribution of depths at a site contains no outliers; only the distribution of depths for Site4 seems very skewed (left-skewed in this case). The IQRs aren't all that far off from one another, so the constant variance across groups (sites in this case) assumption seems reasonable.

Let's check to see that the ratio of the largest group standard deviation to the smallest one is less than 2.

```{r}
sds<-with(df,tapply(depth,siteNum,FUN=sd))
max(sds)/min(sds)
```
It's well below 2, so we're good on the constant variance assumption.

To examine normality with each group a bit further, we'll look at an NPP plot of the group residuals. (uses siteMeans from above)

```{r}
grpResids<-df$depth - siteMeans[as.numeric(df$siteNum)]
qqnorm(grpResids,main="NPP plot of group residuals")
qqline(grpResids)
```

So the assumption normality of each group seems quite reasonable.

***c. Single factor ANOVA on depths vs site***

$H_0$ : mean depth of finding is seem across all sites  
$H_A$ : at least one pair of sites contains differing mean depths


```{r}
test<-aov(depth~siteNum,df)
summary(test)
```

So we have a massive F-score and a miniscule P-value ($O(10^{-6})$), so we can comfortably reject the null hypothesis that the mean depth of findings is the same at all archaelogical sites.

***d./e. Using the iris dataset, determine which among sepal width, petal width, sepal length, and petal length would be suitable for one-way ANOVA using Species as the grouping variable. Then perform the one-way ANOVA procedure.***

Let's see if Sepal.Length is suitable for ANOVA w/ Species as grouping variable.

```{r}
nrow(iris) - nrow(iris[complete.cases(iris),])
## so no missing values for any variable of iris dataset
with(iris,boxplot(Sepal.Length~Species,main="Side-by-side boxplot of\n Sepal Length vs Species\n for iris"))
sepalLengthMeans<-with(iris,tapply(Sepal.Length,Species,FUN=mean))
sepalLengthMeans
points(x=seq_along(iris$Species%>%unique),y=sepalLengthMeans,pch=4,cex=1.5)

sepalLengthSDs<-with(iris,tapply(Sepal.Length,Species,FUN=sd))

max(sepalLengthSDs)/min(sepalLengthSDs)

grpResids<-iris$Sepal.Length - sepalLengthMeans[as.numeric(iris$Species)]

qqnorm(grpResids,main="NPP plot of Sepal.Length for iris")
qqline(grpResids)
```

So Sepal.Length seems good for ANOVA.

***Anova for Sepal.Length vs Species***

```{r}
test<-aov(Sepal.Length~Species,data=iris)
summary(test)
```

***So it's safe to conclude that mean sepal lengths aren't the same across the species setosa, versicolor and virginica***

***We've found one variable suitable for ANOVA w/ Species, so it's time to move on to something else.***