---
title: "Hypothesis Testing and Confidence Intervals for Proportions and for the Difference Between 2 Independent Proportions"
output:
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
  pdf_document: default
---

## Exercise 18.3 (source "The Book of R"  by Tillman Davies)  
***An advertisement for a skin cream claims nine out of ten women who use it would recommend it to a friend. A skeptical salesperson in a department store believes the true proportion of women users who'd recommend it, $\pi$, is much smaller than 0.9. She follows up with 89 random customers who had purchased the skin crea and asks if they would recommend it to others, to which 71 answer yes.***

***a. Set up an appropriate pair of hypotheses for this test and determine whether it will be valid to carry out using the normal distribution.***

**$H_0$**: 90% of all women users would recommend skin cream to friend

**$H_A$**: Fewer than 90% of all women users would recommend skin crea to friend

Expected number of successes is 80.1 and failures is 8.9, both of which are greater than or equal to 5. This satisfies the lack of skew condition.

The sampled customers constitute a random sample of all female customers who us the skin cream that are less than 10% of the total population of all female customers. This satisfies the independence condition. 

***b. Compute the test statistic and the p-value and state your conclusion for the test using a significance level of $\alpha$ = 0.1***

```{r}
onePropZTest<-function(pi0,k,n,alpha=0.05,alt="greater"){
  if(!(pi0>=0 && pi0<=1))
    stop("invalid valid for pi0")
  if(!(k>=0 && k<=n))
    stop("either n or k is invalid")
  if(n*pi0<=5 || n*(1-pi0)<=5)
    stop("sampling distr of p_hat is too skewed to proceed")
  if(!(alpha>0.001 && alpha<0.2))
    warning("the significance level of this test is bizarre")
  if(!(alt %in% c("greater","less","two")))
    warning("performing one-sided hypothesis test with alternative hypothesis pi>pi0")
  sErr<-sqrt((pi0 * (1 - pi0))/n)
  pHat<-k/n
  z<-(pHat - pi0)/sErr
  pVal<-{
    if(alt=="greater"){
      pnorm(q=z,lower.tail=F)
    }else if(alt=="less"){
      pnorm(q=z,lower.tail=T)
    }else{##two-tailed test with p-Value that is double that returned by the pt function; the value of lower.tail passed to the pt function is dictated by the direction of deviation of pHat from pi0
      if(z<0)
        2*pnorm(q=z,lower.tail=T)
      else
        2*pnorm(q=z,lower.tail=F)
    }
  }
  reject<-ifelse(pVal<alpha,"REJECT NULL","FAIL TO REJECT NULL")
  cat("z-score:",z,"\n","p-Value:",pVal,"\n","decision:",reject,"\n",sep="")
}

onePropZTest(0.9,71,89,alt="less",alpha=0.1)

```

***c. Using your estimated sample proportion, construct a two-sided 90% confidence interval for the true proportion of women who would recommend the skin cream***

```{r}
onePropZInterval<-function(k,n,conf=0.95){
if(!(k>=0 && k<=n))
    stop("either n or k is invalid")
pHat<-k/n
if(n*pHat<=5 || n*(1-pHat)<=5)
    stop("sampling distr of p_hat is too skewed to proceed")
if(conf<0.80)
    warning("This interval lacks confidence")
#use pHat as unbiased estimator of pi
sErr<-sqrt(((pHat)*(1-pHat))/n)
z_star<-qnorm((1-conf)/2,lower.tail=F)
ME<-z_star*sErr
bds<- pHat+c(-1,1)*ME
return(cat("(",bds[1],",",bds[2],")",sep=""))
}
onePropZInterval(71,89,conf=0.90)
```

***d. Letting $\pi_1$ denote the true proportion of citizens in support of decriminalization in state 1, and $\pi_2$ the same measure in state 2, conduct and conclude a hypothesis test under a significance level of $\alpha$ = 0.05 with reference to the following hypotheses: <br> $H_0$: $\pi_2$ - $\pi_1$ = 0 <br> $H_A$: $\pi_2$ - $\pi_1$ $\neq$ 0.***

```{r}
#This is a function that performs a z-test for the difference between two independent proportions
diffBtwnTwoIndependentPropDZTest<-function(k1,n1,k2,n2,alpha=0.05,alt="greater"){
  if(!(k1>=0 && k1<=n1)){
    stop("either k1 or n1 is invalid")
  }
  if(! (k2>=0 && k2<=n2)){
    stop("either k2 or n2 is invalid")
  }
  pPooled<-(k1+k2)/(n1+n2)
  if(n1*pPooled<=5 || n1*(1 - pPooled)<=5 || n2*pPooled<=5 || n2*(1 - pPooled)<=5){
    stop("sampling distribution of the difference of proportions is too skewed to proceed")
  }
  est<-k1/n1 - k2/n2 #this is the estimate of hte difference in proportions
  sErr<-sqrt((pPooled * (1 - pPooled))*(1/n1 + 1/n2))
  z<-(est - 0)/sErr
  pVal<-{
    if(alt=="greater"){
      pnorm(q=z,lower.tail=F)
    }else if(alt=="less"){
      pnorm(q=z,lower.tail=T)
    }else{##two-tailed test with p-Value that is double that returned by the pt function; the value of lower.tail passed to the pt function is dictated by the direction of deviation of pHat from pi0
      if(z<0)
        2*pnorm(q=z,lower.tail=T)
      else
        2*pnorm(q=z,lower.tail=F)
    }
  }
  reject<-ifelse(pVal<alpha,"REJECT NULL","FAIL TO REJECT NULL")
  cat("z-score:",z,"\n","p-Value:",pVal,"\n","decision:",reject,"\n",sep="")
}
diffBtwnTwoIndependentPropDZTest(97,445,90,419,alt="two")
```

***e. Compute and interpret a corresponding CI.***

Note, a 95% confidence interval is computed.

```{r}
zIntervalForDiffBtwn2IndependentProportions<-function(k1,n1,k2,n2,conf=0.95){
  if(!(k1>=0 && k1<=n1)){
    stop("either k1 or n1 is invalid")
  }
  if(! (k2>=0 && k2<=n2)){
    stop("either k2 or n2 is invalid")
  }
  #Separately check success/failure condition for each sample
  if(k1<=5 || n1 - k1<=5||k2<=5|| n2 - k2<=5)
    stop("sampling distr of at least one among (phat1,phat2) is too skewed to proceed")
  pHat1<-k1/n1
  pHat2<-k2/n2
  sErr<-sqrt((pHat1*(1-pHat1))/n1 + (pHat2*(1-pHat2))/n2)
  z_star<-qnorm((1-conf)/2,lower.tail=F)
  est<-pHat1 - pHat2
  ME<- z_star * sErr
  bds<-est+c(-1,1)*ME
  return(cat("(",bds[1],",",bds[2],")",sep=""))
}
zIntervalForDiffBtwn2IndependentProportions(97,445,90,419)
```
